Challenge: Using Planning to Generate Natural-Language Sentences

### 2009-12-10 trunk/Papers/crisp-ff-10/crisp-ff.tex
some initial stuff

>>>>

Things that went wrong with FF:

\begin{enumerate}
\item Enforced hillclimbing: Initial state of example problem has a
  relaxed graph of length one because ``expressed'' and ``subst''
  goals are achieved, and newly introduced violations of ``subst'' and
  ``distractor'' are not noticed. As a consequence, EHC on the second
  state spends all of its time in breadth-first search.
\item Goal agenda: For some reason that I don't understand in detail,
  the goal agenda heuristic can choose that the ``subst'' goal must be
  reached before even the ``expressed'' goal. This leads the planner
  to generating some arbitrary sentence. Because the example grammar
  is so small, this leads to a state from which the ``expressed'' goal
  can no longer be achieved. (A realistic big grammar should be able
  to recover from this by introducing a conjunction or relative clause
  or some such -- but of course, this would still lead to a stupid
  sentence.)
\item No helpful actions in best-first search: After the initial
  failure of the goal agenda, FF fell back to best-first search, but
  didn't use the helpful actions heuristic in this search.
\item Tautology detection in reachability analysis: The old FF
  performed a pairwise comparison of the ground literals in the
  instantiated goal to remove tautologies and obvious contradictions.
  In the sentence generation problems, the instantiated goal can
  contain lots of literals, so this takes a long time.  Removing the
  check helped a lot.  Another way to fix this problem would have been
  to sort the list of ground literals appropriately and then only
  compare neighboring literals.
\item Unnecessary computation of hard action template data structure:
  FF computed a big data structure that was only useful for hard
  action templates, although the sentence generation planning problems
  only contain easy action templates.  Because the universe can be
  relatively big, this again wasted a lot of time.
\end{enumerate}

Other observations:

\begin{enumerate}
\item Subst and distractor atoms have a weird lifecycle: They start as
  false, then get made true at some point, and then false again.
  After they have been made false, they can never become true again.
\end{enumerate}

### 2009-12-10 trunk/Papers/crisp-ff-10/crisp.tex
ICAPS submission for CRISP as challenge problem
some initial stuff

\section{Experiences with FF}
\label{sec:crisp-ff}

>>>>

\section{Sentence generation as planning}
\label{sec:crisp}

\subsection{Sentence generation with tree-adjoining grammars}

We consider a planning domain that comes up in the generation of
natural language sentences. This is a standard problem in
computational linguistics and can be construed as follows.

\begin{figure}
  \centering
  \includegraphics[width=0.75\columnwidth]{pic-grammar}
  \caption{An example grammar in the sentence generation domain.}
  \label{fig:white-rabbit-sleeps-grammar}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.75\columnwidth]{pic-derivation}
  \caption{Derivation of ``The white rabbit sleeps.''}
  \label{fig:white-rabbit-sleeps-deriv}
\end{figure}

We assume a tree-adjoining grammar (TAG; \cite{joshi;etal1997}), which
consists of a finite set of \emph{elementary trees} as shown in
Fig.~\ref{fig:white-rabbit-sleeps-grammar}. Elementary trees can be
combined using the operations of \emph{substitution} and
\emph{adjunction}, which are illustrated in
Fig.~\ref{fig:white-rabbit-sleeps-deriv}. Substitution replaces a
substitution node (marked with a downarrow) by a subtree whose root
has the same label, whereas adjunction splits a node in two and
replaces it with a subtree whose root and foot node (marked with an
asterisk) have the same label. A TAG derivation starts with a single
substitution node labeled with a start symbol (in the example, S) and
terminates when all leaves of the tree are labeled with words. In this
way, a TAG grammar defines a language of trees, from which we can read
off a language of strings.

To model the sentence generation problem, we equip every elementary
tree with a set of atoms that specify the \emph{semantic
  representation} of the tree \cite{Stone2003a}.  These terms in these
atoms may be \emph{semantic roles}, such as $\mathsf{subj}$ and
$\mathsf{obj}$ indicating the subject and object positions of a verb;
all substitution nodes of an elementary tree must and the other nodes
may be annotated with semantic roles as well.  In an \emph{instance}
of an elementary tree, all semantic roles have been substituted with
constants from the knowledge base.  Now we can assume a TAG grammar
and two finite sets of ground atoms for the knowledge base and the
semantic representation we want to express in a sentence.  The
sentence generation problem is then to compute a correct TAG
derivation using elementary tree instances such that the input
semantic representation is covered by the semantic representations of
the tree instances; that is, the derivation expresses the intended
meaning.  Furthermore, the assignment of constants to the semantic
roles of the elementary trees must be \emph{unique}: If we replace
each semantic role of every elementary tree in the derivation by a
different variable, there must be exactly one variable assignment that
makes the combined semantic representations of the elementary trees a
subset of the knowledge base.  This ensures that the hearer can
uniquely understand each referring expression.

For example, assume the grammar in
Fig.~\ref{fig:white-rabbit-sleeps-grammar} and a knowledge base that
contains the atoms $\{\mathsf{sleep}(e,r_1), \mathsf{rabbit}(r_1),
\mathsf{rabbit}(r_2), \mathsf{white}(r_1), \mathsf{rabbit}(r_2)\}$,
and let's say we want to express the semantic representation
$\{\mathsf{sleep}(e,r_1)\}$.  Then the derivation in
Fig.~\ref{fig:white-rabbit-sleeps-deriv} would be a correct output of
a sentence generator.  Notice that a derivation for the sentence ``the
rabbit sleeps'' would not be a correct result because the expression
``the rabbit'' is not a unique description of $r_1$: the set
$\{\mathsf{sleep}(x,y), \mathsf{rabbit}(y)\}$ can also be mapped into
the knowledge base by substituting $r_2$ for $y$, so the hearer might
misunderstand the subject as $r_2$. We call $r_2$ a \emph{distractor}
of the referring expression.  Note also that the input semantic
representation did not specify that we should refer to $r_1$ as
``white'' or ``rabbit''; this decision was made by the generation
algorithm.

\subsection{Sentence generation as a planning problem}

The TAG sentence generation problem can be encoded as a planning
problem \cite{KolSto07}.  The key idea is that each operator encodes
the addition of an elementary tree to the derivation; the syntactic
and semantic requirements and effects of doing this are captured in
the operator's precondition and effect.

\begin{figure}
\centering
\begin{minipage}{0.8\textwidth}
{\small%
\begin{verbatim}
(:action sleeps
   :parameters (?u - node  ?subjnode - node  ?nextnode - node
                ?xself - individual  ?xsubj - individual)
   :precondition
       (and (subst S ?u)  (referent ?u ?xself)  (sleep ?xself ?xsubj)
            (current ?subjnode) (next ?subjnode ?nextnode)
   :effect 
       (and (not (subst S ?u))  (expressed sleep ?xself ?xsubj)
            (subst NP ?subjnode)  (referent ?subjnode ?xsubj)
            (not (next ?subjnode)) (next ?nextnode)
            (forall (?y - individual)
                (when (not (= ?y ?xself)) (distractor ?subjnode ?y)))))

(:action rabbit
   :parameters (?u - node  ?xself - individual)
   :precondition 
       (and (subst NP ?u)  (referent ?u ?xself)  (rabbit ?xself))
   :effect 
       (and (not (subst NP ?u))  (canadjoin N ?u)
            (forall (?y - individual)
                (when (not (rabbit ?y)) (not (distractor ?u ?y))))))

(:action white
   :parameters (?u - node  ?xself - individual)
   :precondition 
       (and (canadjoin N ?u)  (referent ?u ?xself)  (rabbit ?xself))
   :effect 
       (forall (?y - individual)
           (when (not (white ?y)) (not (distractor ?u ?y)))))
\end{verbatim}}%
\end{minipage}
\caption{PDDL actions for generating the sentence ``The white rabbit
sleeps.''}
\label{fig:white-rabbit-as-planning}
\end{figure}

More precisely, each operator has a parameter \verb|?u| representing
the syntax node into which the elementary tree is substituted or
adjoined and a parameter \verb|?xR| for each semantic role \verb|R| of
the elementary tree which will be instantiated with the constants with
which the role gets instantiated in the derivation. Furthermore, if
the tree has substitution nodes, there are parameters \verb|?Rnode|
for each role representing these nodes, and an extra parameter
\verb|?nextnode|. The state encodes a list of possible node names;
$\mathsf{current}(n_i)$ expresses that $n_i$ is the next node name
that should be picked when a new substitution node is created, and
$\mathsf{next}(n_i,n_{i+1})$ says that the next node after $n_i$ is
$n_{i+1}$. The atom $\mathsf{subst}(A,u)$ expresses that we need to
still substitute something into the substitution node $u$ with label
$A$; the initial state contains an atom
$\mathsf{subst}(A,\mathsf{root})$ where $\mathsf{root}$ is an
arbitrary name for the root of the derivation.  The values of the
semantic roles are maintained using the $\mathsf{referent}$ atoms.
Finally, we keep track of the uniqueness of referring expressions in
the $\mathsf{distractor}$ atoms: $\mathsf{distractor}(u,a)$ expresses
that the expression at syntax node $u$ could still be misunderstood as
$a$.

The example generation problem above translates into a planning
problem whose domain is shown, in PDDL format, in
Fig.~\ref{fig:white-rabbit-as-planning}. The initial state of the
problem encodes the generation problem's knowledge base; it contains
atoms $\mathsf{rabbit}(r_1)$, $\mathsf{sleep}(e,r_1)$, etc.
Furthermore, the initial state expresses the beginning of the
derivation in atoms $\mathsf{subst}(S,\mathsf{root})$ and
$\mathsf{referent}(\mathsf{root},e)$.  The goal requires syntactic
completeness as $\forall u \forall A \neg \mathsf{subst}(A,u)$ and
unique reference as $\forall u \forall x \neg
\mathsf{distractor}(u,x)$; it also specifies the semantic
representation we want to express in an atom
$\mathsf{expressed}(\mathsf{sleep},e,r_1)$.  A minimal plan for this
problem looks as follows:

\begin{enumerate}
\item $\mathsf{sleeps}(\mathsf{root}, n_1, n_2, e, r_1)$
\item $\mathsf{rabbit}(n_1, r_1)$
\item $\mathsf{white}(n_1, r_1)$
\end{enumerate}

This plan can be automatically decoded into the derivation shown in
Fig.~\ref{fig:white-rabbit-sleeps-deriv}.  The analogous plan with
$r_2$ instead of $r_1$ is not correct because the $\mathsf{sleep}$
precondition of the $\mathsf{sleeps}$ operator is not satisfied.  The
first two steps of the plan alone would not be a correct plan because
they leave an atom $\mathsf{distractor}(n_1, r_2)$ in the state, which
contradicts the goal that all distractors have been eliminated.
